---
title: "chapter6"
author: "Anne Paakinaho"
date: "30 11 2020"
output: html_document
---
## RStudio exercise 6: Analysis of longitudinal data
```{r}
date()
```

Let's download dataset BPRSL and RATSL into R markdown.
These file were made in meet_and_repeat r script. Datasets are in long form.
In BPRS, there is two treatment groups and in total 20 participants in both treatment groups so there are in total of 40 participants (all are male, as said in Kimmo's book). Brief psychiatric rating scale is measured before treatment and during following 8 weeks.

RATS include data from nutrition study; there are three distinct groups of rats that had different diets. Other variables, such as WD1 or WD22 describe the day when the body weight was measured (in grams)

```{r}
BPRSL <- read.table("C:/Users/silve/OneDrive - University of Eastern Finland/IODS-project 2020/IODS-project/data/BPRSL.txt")

RATSL <- read.table("C:/Users/silve/OneDrive - University of Eastern Finland/IODS-project 2020/IODS-project/data/RATSL.txt")

dim(BPRSL)
dim(RATSL)
str(BPRSL)
str(RATSL)
summary(BPRSL)
summary(RATSL)
```
Categorical variables need to be converted into factors again.

```{r}
BPRSL$treatment <- factor(BPRSL$treatment)
BPRSL$subject <- factor(BPRSL$subject)
str(BPRSL)
RATSL$ID <- factor(RATSL$ID)
RATSL$Group <- factor(RATSL$Group)
str(RATSL)
```
Let's do some plotting
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
#Plotting BPRSL
ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))

ggplot(BPRSL, aes(x = week, y = bprs, group = treatment)) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "week", breaks = seq(0, 20)) +
  scale_y_continuous(name = "bprs") +
  theme(legend.position = "top")

```

From the plot, you can see that the BPRS (brief psychiatric rating scale) is higher in the treatment 2 and the BPRS is staying higher through out the data. But in the treatment 1, BPRS are getting smaller by the week go by.
The lower graph seems to be quite useless...

```{r}
#plotting RATSL
ggplot(RATSL, aes(x = Time, y = Weight, group = ID)) +
  geom_line(aes(linetype = Group)) +
  scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 10)) +
  scale_y_continuous(name = "Weight (grams)") +
  theme(legend.position = "top")

ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ ID, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))

```

Most of the rats weigh less than 300 grams, but some individuals are heavier. The heavy rats seem to be in groups 2 and 3.
Rats also seem to get heavier by the day go by.


I will now try to do some analyses on *RATSL* data using chapter 8 of MABS (Kimmo's book) so I will follow datacamp but instead of using BPRSL data I use RATSL.


```{r}
colnames(RATSL)
RATSL$Time
hist(RATSL$Weight)

#standardization
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate(stdweight = (Weight - mean(Weight))/sd(Weight) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSL)

# Plot again with the standardized weight
ggplot(RATSL, aes(x = Time, y = stdweight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ ID, labeller = label_both) +
  scale_y_continuous(name = "standardized weight")

ggplot(RATSL, aes(x = Time, y = stdweight, group = ID)) +
  geom_line(aes(linetype = Group)) +
  scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 10)) +
  scale_y_continuous(name = "stdweight (grams)") +
  theme(legend.position = "top")

```
  
The lower plot is more informative for RATSL data. You can see here the same trend as before, group 2 and 3 contain heavy rats.

Summary graph for RATSL.
```{r}
# Number of measurements in different timepoints
n <- RATSL$Time %>% unique() %>% length()
n #11

# Summary data with mean and standard error of weight by group and time 
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(Weight), se = sd(Weight)/sqrt(n) )%>%
  ungroup()

# Glimpse the data
glimpse(RATSS)

# Plot the mean profiles
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.8,0.8)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
```

The same trend can be seen here in means that rats in group 1 are skinny and in groups 3 the heaviest. Standard errors for group 1 are very small, and for group 2 they seem to be largest.
In this graph, you can also see that there are two recording taken in week 7.


```{r}
# Create a summary data by groups and ID with mean as the summary variable.
RATSS2 <- RATSL %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()
# Glimpse the data
glimpse(RATSS2)

# Draw a boxplot of the mean versus group
ggplot(RATSS2, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight)")
summary(RATSS2$mean)
```

It seems that there is one outlier in each group.
I guess I should filter outliers out. Min values is 237.6 and max 590.5. I will proceed as it was done in datacamp.

```{r}
RATSS21 <- RATSS2 %>%
  filter(mean >238 & mean < 589)

ggplot(RATSS21, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight)")
str(RATSS21)
```

The outliers in group 1 and group 2 are gone, but I don't know how to deal with the outlier in group 3.
 
I would like to do t-test. But there are three groups...
I check out what analysis of variance gives me... ANOVA can be used if there is 3 or more variables.
```{r}
# Fit the linear model with the mean as the response
fit <- lm(mean ~ Group, data = RATSS21)
anova(fit)
```

Groups are significantly different from each other (P<0.05). 

I will continue now with *BPRSL* data.
I will fit a multiple linear regression model with bprs as response and week and treatment as explanatory variables.
```{r}
colnames(BPRSL)
hist(BPRSL$bprs)
# create a regression model BPRS_reg
BPRS_reg <- lm(bprs ~ week + treatment, data = BPRSL)
summary(BPRS_reg)
```

In histogram, bprs is not really normally distributed, it is more skewed to the right.
This linear regression was a fixed model.  
Residuals are not symemtrical..
Measurement timepoint, week, is significantly related to BPRS, but treatment is not.
Multiple R-squared is 0.1851 so together week and treatment account less than 20% of the variation in BPRS.


Random intercept model (random model as the name says).
```{r}
# access library lme4
library(lme4)
# Create a random intercept model
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)

# Print the summary of the model
summary(BPRS_ref)
```
  
In random intercept model it is possible that the intercept can differ for each subject in linear regression fit.
Scaled residuals seem pretty symmetrical.
The variance for random effects is 47,41, I guess it is somewhat large, so there is variation in intercepts of the regression fits of individual subjects.


Random Intercept and Random Slope Model; it allows the linear regression fits for each subject to differ in slope.
```{r}
# create a random intercept and random slope model
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref1)
# perform an ANOVA test on the two models
anova(BPRS_ref1, BPRS_ref)
```

In Random Intercept and Random Slope Model, residuals are quite symmetrical. Variance is larger than in Random intercept model. 
ANOVA: The difference between these two models is significantly different (P<0.05).


Random Intercept and Random Slope Model with interaction:week and treatment interaction
````{r}
BPRS_ref2 <- lmer(bprs ~ week * treatment + (week | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref2)
anova(BPRS_ref2, BPRS_ref1)
````

AIC values get smaller all the time when changing the model, but only little. But the smaller, the better.
BPRS_ref2 and BPRS_ref2 are no longer different from each other, P>0.05. 

Let's find the fitted values from the interaction model and plot the fitted values for each subject:
```{r}
ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))

# Create a vector of the fitted values
Fitted <- fitted(BPRS_ref2)

# Create a new column fitted to RATSL
BPRSL <- BPRSL %>%
  mutate(Fitted)

ggplot(BPRSL, aes(x = week, y = Fitted, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))
````

Fitted plot follows the same pattern as the true values for subjects.

*The End, hurray!*
